---
title: "Random Variables and Plots"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Random Variables and Plots

This file will show how to generate random samples and filter data, and how to plot data. 


```{r}
dat <- read.csv("femaleMiceWeights.csv")
install.packages("downloader") 
library(downloader)
head(dat)
dat[12,2]#12 row, second column
col_11 <- (dat$Bodyweight)[11]
col_11 
head(dat)
nrow(dat)
max(dat$Bodyweight)
row37 <- dat[3:7,]
row37
mean(row37$Bodyweight)
```

# Generate a random sample of size 1 from the numbers 13 to 24

```{r}
random_row <- sample(13:24, size = 1)
```

# Find the weight of the mouse represented by that row
```{r}
mouse_weight <- dat$Bodyweight[random_row]
mouse_weight
```

# select specific column

Use filter and select 

```{r}
install.packages("dplyr")
install.packages("utf8")
library(dplyr)
controls <- filter(dat, Diet=="chow")
controls <- select(controls, Bodyweight)
unlist(controls)
controls <- filter(dat,Diet=="chow") %>% 
  select(Bodyweight) %>% unlist
```


# Filter rows where "order" is equal to "primates"

```{r}
library(dplyr)
filtered_msleep <- msleep %>%
  filter(order == "Primates")
```


# View the resulting data frame

```{r}
View(filtered_msleep)
nrow(filtered_msleep)
class(filtered_msleep)
```

# Plots 

seq:it generates values
It takes three arguments: the start value, the end value, and the step size (increment)
floor(min(x):largest integer less than or equal to the minimum value in the dataset 'x.'
ceiling(max(x):  smallest integer greater than or equal to the maximum value in the dataset 'x.'

# Histogram
```{r}
install.packages("UsingR")
library(UsingR)
install.packages("htmlTable")
library(UsingR)
x=father.son$fheight
length(x)
round(sample(x,20),1)

hist(x,breaks=seq(floor(min(x)),ceiling(max(x))),
     main="Height histogram",xlab="Height in inches")
```

# line plot
ecdf:empirical cumulative probability distribution of 'x'
(xs) is used to evaluate the ECDF at the specified sequence of 'xs.'
l:line plot 
```{r}
xs<-seq(floor(min(x)), ceiling(max(x)),0.1)
plot(xs,ecdf(x)(xs),type="l",
     xlab="Height in inches",ylab="F(x)")
```


## QQ Plot

We can tell if data is normally distributed from QQ Plot. 

```{r}
ps <- seq(0.01,0.99,.01)
qs <- quantile(x,ps)
normalqs <-pnorm(ps,mean(x),sd(x))
plot(normalqs,qs,xlab="Normal percentiles",ylab="Height percentiles")
abline(0,1) 

load("skew.RData")
dim(dat)#dimension
par(mfrow = c(3,3))
for (i in 1:9) {
  qqnorm(dat[, i], main = paste("QQ Plot - Variable", i))
}
```

## Box Plot
when data is not normally distributed , QQ is not effective. Boxplot is good. 

```{r}
data(exec.pay, package="UsingR")
hist(exec.pay)
qqnorm(exec.pay)
qqline(exec.pay)

boxplot(exec.pay, ylab="10,000s of dollars", ylim=c(0,400))
```

Box Plot using split

```{r}
head(InsectSprays)
boxplot(split(InsectSprays$count, InsectSprays$spray), main = "Boxplot by Spray", xlab = "Spray", ylab = "Count")
```

Box Plot using formula

```{r}
boxplot(count ~ spray, data = InsectSprays, main = "Boxplot by Spray", xlab = "Spray", ylab = "Count")
```


## Compare between two mice group (healthy and overweight)

