---
title: "Statistical Tests"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Statistical Tests

This file has concepts about Null hypothesis , p-value, t-test, and CLT. 


## Compare between two mice group (healthy and overweight)

```{r }
dat <- read.csv("femaleMiceWeights.csv")
control <- filter(dat,Diet=="chow") %>% 
  select(Bodyweight) %>% unlist
treatment <- filter(dat, Diet=="hf") %>%
  select(Bodyweight) %>% unlist
mean(treatment)
mean(control)
```

Now will access the full population and take a sample to to help us answer if the difference happened by chance or not 
```{r }
population <-read.csv("femaleControlsPopulation.csv")
population <-unlist(population)
mean(sample(population,12))
```


# Excercises
```{r }
library(downloader)
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleControlsPopulation.csv"
filename <- basename(url)
download(url, destfile=filename)
x <- unlist(read.csv(filename)) #entire population
RNGkind("Mersenne-Twister", "Inversion", "Rejection") 
mean(x)
set.seed(5)
x2 <- sample(population,5)
abs(mean(x2) - mean(x))
```

## Null hypothesis and p-value

If p-value <0.05 : there is a significant difference between the two groups

```{r }
obs <- mean(treatment) - mean(control)
n <- 10000
nulls <- vector("numeric",n)#empty vector
for (i in 1:n){
  control <- sample(population,12)
  treatment <-sample(population,12)
  nulls[i] <- mean(treatment) - mean(control)
}
hist(nulls)
mean(abs (nulls) > obs)
```

## Use Central Limit Theorem

The theorem states that if you take a large sample from a population, the sample's mean will be normally distributed. 
```{r }
library(rafalib)
qqnorm(nulls)
qqline(nulls)
```

## Exercises
```{r }
set.seed(1)
n <-1000
sampleavg <- vector("numeric",n)#empty vector
for (i in 1:n){
  sampledat <- sample (x, 5)
  sampleavg[i] <- mean(sampledat)
}
prop <- mean((abs(sampleavg) - mean(x)) > 1)
```

## Probability Distributions 
```{r }
install.packages("gapminder")
library(gapminder)
data(gapminder)
head(gapminder)
```

What is the proportion of countries in 1952 that have a life expectancy less than or equal to 40?
```{r }
lifex <- filter(gapminder, year=="1952") %>%
  select(lifeExp) %>% unlist
mean(lifex <=40)
```

## Normal Distribution
mu and sigma are the most important values
when we standardize data (find z), will have have same units. 
```{r }
x <= unlist (read.csv("femaleControlsPopulation.csv"))
set.seed(1)
n <- 1000
averages5 <- vector ("numeric",n)
for (i in 1:n){
  X <- sample(x,5)
  averages5[i] <- mean(X)
```

```{r }
set.seed(1)
n <- 1000
averages50 <- vector("numeric",n)
for (i in 1:n) {
  X <- sample(x,50)
  averages50[i] <- mean(X)
}
hist(averages5)
hist(averages50)
```

Calculate the proportion between 23 and 25 
```{r }
proportion_between_23_and_25 <- mean(averages50 > 23 & averages50 < 25)
cat("Proportion of averages between 23 and 25:", proportion_between_23_and_25, "\n")
```

What is the proportion of observations between 23 and 25 in a normal distribution with average 23.9 and standard deviation 0.43?
```{r }
mu <- 23.9
sigma <- 0.43
prop25 <- pnorm(25, mean=mu, sd= sigma)
prop23 <- pnorm(23, mean=mu, sd= sigma)
prop2325 <- prop25 - prop23 
cat("Proportion of averages between 23 and 25 in normal distribution  :", prop2325, "\n")
```

## Population and Samples CLT

```{r }
dat <- read.csv("mice_pheno.csv")
#remove the lines that contain missing values
dat <- na.omit(dat)
library(dplyr)
x <- filter(dat, Diet=="chow" & Sex == "M") %>% select(Bodyweight) %>% unlist
mean(x)
sd(x)
set.seed(1)
X = sample(x,25)
X_mean <-  mean(X)
#Use dplyr to create a vector y with the body weight of all males on the high fat hf) diet.
#What is this population's average?
y = filter(dat,Diet=="hf" & Sex == "M" ) %>% select(Bodyweight) %>% unlist
mean(y)
popsd(y)
```

Use dplyr to create a vector y with the body weight of all males on the high fat hf) diet.#What is this population's average?
```{r }
y = filter(dat,Diet=="hf" & Sex == "M" ) %>% select(Bodyweight) %>% unlist
mean(y)
popsd(y)
```

Set the seed at 1. Take a random sample  of size 25 from y.
```{r }
set.seed(1)
Y = sample(y,25)
Y_mean <- mean(Y)
abs(Y_mean - X_mean)
abs( ( mean(y) - mean(x) ) - ( mean(Y) - mean(X) ) )

x <- filter(dat, Sex=="F" & Diet=="chow") %>% select(Bodyweight) %>% unlist
set.seed(2)
X <- sample(x,25)
y <- filter(dat, Sex=="F" & Diet=="hf") %>% select(Bodyweight) %>% unlist
set.seed(2)
Y <- sample(y,25)
abs( ( mean(y) - mean(x) ) - ( mean(Y) - mean(X) ) )
```

If a list of numbers has a distribution that is well approximated by the normal distribution,
what proportion of these numbers are within one standard deviation away from the list's average?
```{r }
list_of_numbers <- rnorm(100, mean = 0, sd = 1)
mean_val <- mean(list_of_numbers)
sd_val <- sd(list_of_numbers)
cum_prob_upper <- pnorm(mean_val + sd_val, mean = mean_val, sd = sd_val)
cum_prob_lower <- pnorm(mean_val - sd_val, mean = mean_val, sd = sd_val)
```

Calculate the proportion within one standard deviation?
```{r }
proportion_within_one_sd <- cum_prob_upper - cum_prob_lower
cat("Proportion within one standard deviation:", proportion_within_one_sd, "\n")
```

Calculate the proportion within 2 standard deviation?
```{r }
prob_upper <- pnorm(mean_val + 2*sd_val, mean = mean_val, sd = sd_val)
prob_lower <- pnorm(mean_val - 2*sd_val, mean = mean_val, sd = sd_val)
proportion <- prob_upper - prob_lower
```

Calculate the proportion within 3 standard deviation
```{r }
pnorm(3) - pnorm(-3)
```

Define y to be the weights of males on the control diet.
What proportion of the mice are within one standard deviation away from the average weight?
```{r }
y <- filter(dat, Sex=="M" & Diet == 'chow') %>% select(Bodyweight) %>% unlist
library(rafalib)
mean_val <- mean(y)
sd_val <-popsd(y)
```

Calculate the proportion within one standard deviation?
```{r }
z_score <- (y - mean(y))/popsd(y) 
mean(abs(z_score) <= 1)
```


## T-test
```{r }
dat <- read.csv("femaleMiceWeights.csv")
#Healthy group
control <- filter(dat,Diet=="chow") %>% 
  select(Bodyweight) %>% unlist
#overweight group
treatment <- filter(dat, Diet=="hf") %>%
  select(Bodyweight) %>% unlist
N <- length(treatment)
obs <- mean(treatment) - mean(control)
```

to do t-test we need to estimate the standard error
```{r }
se <- sqrt(
  var(treatment)/N +
    var(control)/N
)
```

obtain t-value
```{r }
ttest <- obs/se
```

P-value for one tail
```{r }
1 - pnorm(ttest)
```

P-value for two tail
```{r }
2* (1-pnorm(ttest))
```

t-test
```{r }
ttest <- t.test(treatment, control)
ttest
qqnorm(control)
qqline(control)
```

## Calculating z-score
n: Number of dice rolls
p : probability

```{r }
set.seed(1)
n <- 100 
n_simulations <- 10000
p <- 1/6
simulate_z <- function() {
  x <- sample(1:6, size = n, replace = TRUE)
  proportion_6s <- mean(x == 6)  
  z <- (proportion_6s - p) / sqrt(p * (1 - p) / n)
  return(z)
}
```

```{r }
simulations <- replicate(n_simulations, simulate_z())
proportion_greater_than_2 <- mean(abs(simulations) > 2)
cat("Proportion of times |z| was larger than 2:", proportion_greater_than_2, "\n")
qqnorm(simulations)
qqline(simulations)
```

## Compare between 4 different ps and ns ans see which one has the best normal Distribution
```{r }
ps <- c(0.5,0.5,0.01,0.01)
ns <- c(5,30,30,100)
library(rafalib)
mypar(4,2)
for(i in 1:4){
  p <- ps[i]
  sides <- 1/p
  n <- ns[i]
  zs <- replicate(10000,{
    x <- sample(1:sides,n,replace=TRUE)
    (mean(x==1) - p) / sqrt(p*(1-p)/n)
  }) 
  hist(zs,nclass=7)
  qqnorm(zs)
  abline(0,1)
}
```


## CLT
```{r }
dat <- read.csv('femaleMiceWeights.csv')
X <- filter(dat, Diet=="chow") %>% select(Bodyweight) %>% unlist
Y <- filter(dat, Diet=="hf") %>% select(Bodyweight) %>% unlist
```

# Use the CLT to approximate the probability that our estimate X is off by more than 2 grams from mu 

se : sample mean standard error

```{r }
sigma <- sd(X) 
n <- length(X)
sample_avg <- mean(X)
se <- sigma/ sqrt(n) 
```

# Calculate the z-scores for 2 grams above and below the mean
```{r }
z_upper <- (sample_avg + 2 - mean(X)) / se
z_lower <- (sample_avg - 2 - mean(X)) / se
```

# Use the pnorm function to calculate the probabilities
```{r }
prop_u <- 1- pnorm(z_upper)
prop_l <- pnorm(z_lower)
```

# Sum the probabilities for more than 2 grams above and below
```{r }
sum <- prop_u + prop_l
cat("Probability that the estimate is off by more than 2 grams from the true mean:",sum, "\n")
```

## Quiz week 2
## Question 1
```{r }
RNGkind("Mersenne-Twister", "Inversion", "Rejection")
library(downloader) 
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleControlsPopulation.csv"
filename <- basename(url)
download(url, destfile=filename)
x <- unlist( read.csv(filename) )
```

# What proportion of these 1,000 averages are more than 1 gram away from the average of x ?
```{r }
set.seed(1)
n <- 1000
res <- vector('double',n)
for (i in seq(n)) {
  avg_sample <- mean(sample(x,50))
  res[[i]] <- avg_sample
}
mean(abs(res-mean(x)) > 1)
```

## Question 2
#What is the proportion of countries in 1952 that have a life expectancy longer than 40 years but shorter than 60 years?

```{r }
library(gapminder)
data(gapminder)
head(gapminder)
life_ex = filter(gapminder, year=="1952") %>% select(lifeExp) %>% unlist
life_x = sum(life_ex >40 & life_ex<60)
nnrow = nrow(filter(gapminder, year == 1952))
life_x/nnrow
```

